---
title: "Gov 50 Final Project"
author: "Conner Huey"
description: "My Final Project"
output:
  distill::distill_article:
    self_contained: false
---

I have produced six data visualizations pertaining to my research question of the effects of GDP per capita on democratization. These visualizations provide insight into differences in the effects of GDP per capita on democratization based off continent. My graphs for the Western World, Latin America, and Sub-Saharan Africa appear to demonstrate a generally positive relationship between GDP per capita and democratization. For Eastern Europe, this relationship appears to be positive until per capita GDP reaches $4.5 thousand before turning into a negative relationship. For Asian countries, this relationship appears ot be positive until per capita GDP reaches $9 thousand before turning into a negative relationship. And for North African/Middle Eastern countries, this relationship appears to be negative until per capita GDP reaches approximately $35 thousand before turning into a positive relationship. These findings appear to demonstrate that depending on continent, the effects of GDP per capita on democratization may vary.

## Data Analysis
```{r}
library(tidyverse)
civilwar <- read_csv("civilwar.csv")

civilwar |>
  filter(western == 1) |>
  ggplot(aes(x = gdpen, y = polity2)) +
  geom_point(size = 0.5) +
  geom_smooth() +
  ylim(-10, 10) +
  labs(
    x = "Per Capita Gross Domestic Product in Thousands of USD",
    y = "Democratization Index",
    title = "Effect of Per Capita Gross Domestic Product on Democratization in Western \n\ Countries and Japan",
    caption = "Note: Dollars are in 1985 Dollars"
  )

civilwar |>
  filter(eeurop == 1) |>
  ggplot(aes(x = gdpen, y = polity2)) +
  geom_point(size = 0.5) +
  geom_smooth() +
  ylim(-10, 10) +
  labs(
    x = "Per Capita Gross Domestic Product in Thousands of USD",
    y = "Democratization Index",
    title = "Effect of Per Capita Gross Domestic Product on Democratization in Eastern Europe",
    caption = "Note: Dollars are in 1985 Dollars"
  )

civilwar |>
  filter(lamerica == 1) |>
  ggplot(aes(x = gdpen, y = polity2)) +
  geom_point(size = 0.5) +
  geom_smooth() +
  ylim(-10, 10) +
  labs(
    x = "Per Capita Gross Domestic Product in Thousands of USD",
    y = "Democratization Index",
    title = "Effect of Per Capita Gross Domestic Product on Democratization in Latin America",
    caption = "Note: Dollars are in 1985 Dollars"
  )

civilwar |>
  filter(ssafrica == 1) |>
  ggplot(aes(x = gdpen, y = polity2)) +
  geom_point(size = 0.5) +
  geom_smooth() +
  ylim(-10, 10) +
  labs(
    x = "Per Capita Gross Domestic Product in Thousands of USD",
    y = "Democratization Index",
    title = "Effect of Per Capita Gross Domestic Product on Democratization in \n\ Sub-Saharan Africa",
    caption = "Note: Dollars are in 1985 Dollars"
  )

civilwar |>
  filter(asia == 1) |>
  ggplot(aes(x = gdpen, y = polity2)) +
  geom_point(size = 0.5) +
  geom_smooth() +
  ylim(-10, 10) +
  labs(
    x = "Per Capita Gross Domestic Product in Thousands of USD",
    y = "Democratization Index",
    title = "Effect of Per Capita Gross Domestic Product on Democratization in Asia",
    caption = "Note: Dollars are in 1985 Dollars"
  )

civilwar |>
  filter(nafrme == 1) |>
  ggplot(aes(x = gdpen, y = polity2)) +
  geom_point(size = 0.5) +
  geom_smooth() +
  ylim(-10, 10) +
  labs(
    x = "Per Capita Gross Domestic Product in Thousands of USD",
    y = "Democratization Index",
    title = "Effect of Per Capita Gross Domestic Product on Democratization in \n\ North Africa/Middle East",
    caption = "Note: Dollars are in 1985 Dollars"
  )
```


## Data Visualization

I have produced two data visualizations pertaining to my research question of the effects of GDP per capita on a country's democratization. I find an interesting difference between my two visualizations. My first visualization explores the effects of GDP per capita on countries' democratization in 1999, finding that there seems to exist a positive relationship between GDP per capita and a country's democratization. However, my second visualization, which explores this relationship between GDP per capita and democratization across all countries from 1945 to 1999, there appears to be a positive relationship between GDP per capita and democratization until approximately $10,000 USD per capita when the relationship becomes negative, implying that as per capita GDP increases beyond $10,000, the more likely a country is authoritarian.

```{r}
library(tidyverse)
civilwar <- read_csv("civilwar.csv")

civilwar |>
  filter(year == 1999) |>
  ggplot(aes(x = gdpen, y = polity2)) +
  geom_point(size = 0.5) +
  geom_smooth() +
  ylim(-10, 10) +
  labs(
    x = "Per Capita Gross Domestic Product in Thousands of USD",
    y = "Democratization Index",
    title = "Effect of Per Capita Gross Domestic Product on Democratization in 1999",
    caption = "Note: Dollars are in 1985 Dollars"
  )
```
```{r}
civilwar |>
  ggplot(aes(x = gdpen, y = polity2)) +
  geom_point(size = 0.5) +
  geom_smooth() +
  ylim(-10, 10) +
  labs(
    x = "Per Capita Gross Domestic Product in Thousands of USD",
    y = "Democratization Index",
    title = "Effect of Per Capita Gross Domestic Product on Democratization",
    caption = "Note: Dollars are in 1985 Dollars"
  )
```

## Project Proposal

The research question for my final project is "what are the effects of gross domestic product on a country's democratization index?" I hypothesize that the higher a country's GDP, the more likely a country is to have a higher democracy index score. My hypothesis is rooted in the literature of comparative politics and specifically the modernization theory of democratization which argues that there exists a strong relationship between economic wealth and democracy. For instance, 19 out of the 20 richest countries in the world (aside from microstates and oil kingdoms) are democracies (Singapore is the exception), while the world's 20 poorest countries are authoritarian regimes. Put simply by political scientist Seymour Lipset, "the more well-to-do a nation, the greater the chances that it will sustain democracy" (1959). 

My explanatory variable of interest is the GDP per capita of a country, which is coded as "gdpen." It is recorded in tens of thousands of dollars and is in 1985 dollars. My outcome variable of interest is a country's polity score, which is coded on a scale from -10 (least democratic/most authoritarian) to 10 (most democratic/least authoritarian). If my hypothesis is correct, we should expect to see a positive relationship between GDP per capita and polity score. If my hypothesis is incorrect, we will not see a relationship between the GDP per capita and polity score.


## Project thoughts

For my final project, I'm interested in exploring topics broadly related to voter identity and political behavior from the 1990s to the present day. In particular, I'm interested in exploring either voter shift during the 1994 Congressional election, which coincides with Newt Gingrich's Contract with America, or the 2016 presidential election.